MSAD_PCA_Walkthrough

### 2017-07-13 LaTeX_Vignette/MS-MVSPC_DissertationVignette.Rnw
Finished draft 2 edits.

\subsection{Synthetic Fault Induction}
Faults can be introduced to single- or multi-state data via the \texttt{mspProcessData} function. The default fault start index is 8500, or roughly 84\% through the cycle of 10,080 observations. These faults are added through the internal \texttt{faultSwitch} function.
\begin{enumerate}
\item Fault 1A is a positive shift to all three features before state rotation: $\textbf{X}^*(t_s) = \textbf{X}(t_s) + 2,\ s \ge 8500$.
%
\item Fault 1B is a positive shift to the $x$ feature before state rotation: $x^*(t_s) = x(t_s) + 2,\ s \ge 8500$.
%
\item Fault 1C is a positive shift to the $x$ and $z$ features in State 3 only and \emph{after} state rotation: $x^*(t_s) = x(t_s) + 2,\ z^*(t_s) = z(t_s) + 2,\ s \ge 8500$.
%
\item Fault 2A is a positive drift across all the process monitoring features before state rotation: $\textbf{X}^*(t_s) = \textbf{X}(t_s) + (s - 8500)\times 10^{-3}$, $s > 8500$.
%
\item Fault 2B is a positive drift across the $y$ and $z$ process monitoring features before state rotation: $y^*(t_s) = y(t_s) + (s - 8500)\times 10^{-3},\ z^*(t_s) = z(t_s) + (s - 8500)\times 10^{-3}$, $s > 8500$.
%
\item Fault 2C is a negative drift in the $y$ process monitoring feature in State 2 only and \emph{after} state rotation: $y^*(t_s) = y(t_s) - 1.5 \times \frac{s - 8500}{10080 - 8500}$, for $s > 8500$.
%
\item Fault 3A is an amplification of the underlying latent variable $t$ for all features. The maximum latent drift of this fault will be 5 + 1: $\textbf{X}^*(t_s) = \textbf{X}(t_s^*),\ s > 8500$, where $t_s^* = \left[\frac{5(s - 8500)}{\omega - 8500} + 1\right]t_s$.
%
\item Fault 3B is a mutation of the underlying latent variable $t$ for the $z$ feature: $z^*(t_s) = z(\log t_s^*)$, $s \ge 8500$. This fault will dampen the underlying latent effect for $z$ if $t_s > 1$ and amplify this effect if $t_s < 1$.
%
\item Fault 3C is a polynomial mutation of the error for the $y$ feature in State 2 only and \emph{after} state rotation: $y^*(t_s) = y(t_s) + 2 * \textbf{e}_3(s) - 0.25$, for $s > 8500$.
\end{enumerate}

@@

This section provides a fully commented code walk-through for the main \texttt{msp*} functions in the \texttt{mvMonitoring} package. This package depends on a few other packages, but most of these dependencies are on commonly-used packages. The exception is \texttt{BMS}.
<<r load_mvMonitoring, warning=FALSE, message=FALSE>>=
library(devtools)
library(xts)
# install.packages("BMS") # If necessary
build("C:/Users/gabriel_odom/Documents/GitHub/mvMonitoring/mvMonitoring")
library(mvMonitoring)
@

\subsection{Generating Synthetic Data}
First begin by generating multi-state data from a fault scenario. This code will yield observations under Fault 2A, as described the Synthetic Fault Induction section. We choose the default options for the period length (7 days * 24 hours * 60 minutes = 10,080 observations), the starting index of the fault (8500 out of 10080), and the time stamp for beginning the data is 16 May of 2016 at 10:00AM (my wedding anniversary). As we can see from the \texttt{str()} function, we have an \texttt{xts} matrix with 10080 rows and four columns (the state indicator and the three features).

@@

\subsection{Train the Fault Detection Threshold}
Now that these observations are generated and stored in memory, the \texttt{mspTrain} function can train the MSAD-PCA model. The last 1620 observations (27 hours' worth) will be saved for testing. The \texttt{mspTrain} function takes in the training data set partitioned between the observation and the label column. If this function errors, make sure the label column is not included in the data matrix -- this will cause a singularity in the data. The function will
\begin{enumerate}
\item Train on the first three days' worth of observations, as set by \texttt{trainObs}.
%
\item Scan the fourth day for anomalies, as set by \texttt{updateFreq}.
%
\item Remove any alarmed observations.
%
\item ``Forget'' the first day's observations.
%
\item ``Learn'' the non-alarmed observations from the fourth day.
%
\item Retrain and repeat until the end of the data matrix
\end{enumerate}

Furthermore, the \texttt{Dynamic = TRUE} option means that the \texttt{mspTrain} function will include the lags specified by the \texttt{lagsIncluded} argument. Finally, the number of sequential anomalous observations necessary to raise an alarm is set at 5 by the \texttt{faultsToTriggerAlarm} argument. These last three arguments are set to their defaults.
<<r train>>=
train1A_xts <- fault1A_xts[1:8461,]
# This function will run in 13 seconds on the author's machine.
train1A_ls <- mspTrain(data = train1A_xts[,-1],
                       labelVector = train1A_xts[,1],
                       trainObs = 3 * 24 * 60,
                       updateFreq = 1 * 24 * 60,
                       Dynamic = TRUE,
                       lagsIncluded = 0:1,
                       faultsToTriggerAlarm = 5)
str(train1A_ls)
@
The \texttt{mspTrain} function returns a list of four objects:
\begin{enumerate}
\item \texttt{FaultChecks}: An \texttt{xts} matrix of monitoring statistics and associated indicators for all observations after the burn-in of \texttt{trainObs}. It will have 10080 - \texttt{trainObs} number of rows and five columns:
	\begin{enumerate}
	\item \texttt{SPE}: the SPE statistic for each observation.
    %
    \item \texttt{SPE\_Flag}: an indicator showing if the SPE statistic for that observation is beyond the calculated threshold; 0 is normal, 1 is flagged.
    %
    \item \texttt{T2}: the $T^2$ statistic for each observation.
    %
    \item \texttt{T2\_Flag}: an indicator showing if the $T^2$ statistic for that observation is beyond the calculated threshold; 0 is normal, 1 is flagged.
    %
    \item \texttt{Alarm}: an indicator showing if the observation is in a sequence of flagged observations; 0 is normal, 1 is alarmed.
	\end{enumerate}
%
\item \texttt{Non\_Alarmed\_Obs}: An \texttt{xts} matrix will all observations with an alarm code of 0 from \texttt{FaultChecks}. Of note, this matrix contains the data, while the \texttt{FaultChecks} matrix only contains the monitoring statistics and indicators.
%
\item \texttt{Alarms}: An \texttt{xts} matrix of all the observations removed from the training data matrix.
%
\item \texttt{TrainingSpecs}: a list with length equal to the number of classes -- in this case 3. For each class, this list contains a list of six objects:
	\begin{enumerate}
	\item \texttt{SPE\_Threshold}: a named numeric scalar of the $1 - \alpha$ percentile of the non-parametric estimate of the SPE statistic density.
    %
    \item \texttt{T2\_Threshold}: a named numeric scalar of the $1 - \alpha$ percentile of the non-parametric estimate of the $T^2$ statistic density.
    %
    \item \texttt{projectionMatrix}: The $p \times q$ matrix of eigenvectors necessary to project a $p$-dimensional observation to $q$ dimensions. This is necessary to reduce the dimension of any test observation, and is used in calculating the SPE statistic for test observations.
    %
    \item \texttt{LambdaInv}: The inverse of the diagonal $q \times q$ matrix of eigenvalues. This matrix is used to calculate the $T^2$ statistic for test observations.
    %
    \item \texttt{muTrain}: The mean vector of the training observations. This is used to center the test observations on the training mean.
    %
    \item \texttt{RootPrecisTrain}: The $p \times p$ diagonal matrix of the inverse square roots of the feature variances. This is used to scale the test observations into the training scale.
	\end{enumerate}
\end{enumerate}

@@

\subsubsection{Monitoring the Test Data}
With the lagged test observations in the working environment, the \texttt{mspMonitor} function can be applied. This function (similarly to \texttt{mspTrain}) takes in the label information as a separate argument from the input data. Further, the \texttt{mspMonitor} function takes in the \texttt{TrainingSpecs} object returned in the results list from \texttt{mspTrain}. Notice that the first six rows of the monitor matrix are the exact same as the first six rows of the lagged test matrix, except that the rows of the monitor matrix have the monitoring statistic and corresponding indicator columns appended.
<<r monitor>>=
monitor1A_xts <- mspMonitor(observations = lagTest1A_xts[,-1],
                            labelVector = lagTest1A_xts[,1],
                            trainingSummary = train1A_ls$TrainingSpecs)
head(monitor1A_xts)
@

\subsection{Warn Operators During Alarms}
Note that all `Alarm` codes in the monitor matrix above are recorded with \texttt{NA} values. This is because the \texttt{mspMonitor} function does not check the sequential flag conditions of the monitor matrix. This is the responsibility of the \texttt{mspWarning} function. Because the \texttt{mspWarning} function is designed to test one incoming observation at a time through a script or batch file, the following example is designed to mimic the behavior of the \texttt{mspWarning} function as each new observation comes online.
<<r warnings>>=
alarm1A_xts <- monitor1A_xts
for(i in 1:nrow(alarm1A_xts)){
  if(i < (5 + 1)){
    alarm1A_xts[1:i,] <- mspWarning(alarm1A_xts[1:i,])
  }else{
    alarm1A_xts[(i - 5):i,] <- mspWarning(alarm1A_xts[(i - 5):i,])
  }
}
@
The fault was introduced at index 8500, which corresponds to about 40 minutes into the test hour.
<<r warning_Plot>>=
plot(alarm1A_xts[, ncol(alarm1A_xts)],
     main = "Alarm Codes for Test Data")
@
The alarm codes are
\begin{enumerate}
\item ``0'': No alarm.
\item ``1'': Hotelling's $T^2$ alarm.
\item ``2'': Squared Prediction Error alarm.
\item ``3'': Both alarms.
\end{enumerate}
As we can see, the monitoring function detects a process anomaly after 40 minutes into the test hour, and the warning function issues the corresponding alarms.

>>>>

\subsection{Synthetic Fault Induction} \label{subsec:D_FaultDefns}
Faults can be introduced to single- or multi-state data via the \texttt{mspProcessData} function. The default fault start index is 8500, or roughly 84\% through the cycle of 10,080 observations. These faults are added through the internal \texttt{faultSwitch} function.
\begin{enumerate}
\item Fault 1A is a positive shift to all three features before state rotation: $\textbf{X}^*(t_s) = \textbf{X}(t_s) + 2,\ s \ge 8500$.
%
\item Fault 1B is a positive shift to the $x$ feature before state rotation: $x^*(t_s) = x(t_s) + 2,\ s \ge 8500$.
%
\item Fault 1C is a positive shift to the $x$ and $z$ features in State 3 only and \emph{after} state rotation: $x^*(t_s) = x(t_s) + 2,\ z^*(t_s) = z(t_s) + 2,\ s \ge 8500$.
%
\item Fault 2A is a positive drift across all the process monitoring features before state rotation: $\textbf{X}^*(t_s) = \textbf{X}(t_s) + (s - 8500)\times 10^{-3}$, $s > 8500$.
%
\item Fault 2B is a positive drift across the $y$ and $z$ process monitoring features before state rotation: $y^*(t_s) = y(t_s) + (s - 8500)\times 10^{-3},\ z^*(t_s) = z(t_s) + (s - 8500)\times 10^{-3}$, $s > 8500$.
%
\item Fault 2C is a negative drift in the $y$ process monitoring feature in State 2 only and \emph{after} state rotation: $y^*(t_s) = y(t_s) - 1.5 \times \frac{s - 8500}{10080 - 8500}$, for $s > 8500$.
%
\item Fault 3A is an amplification of the underlying latent variable $t$ for all features. The maximum latent drift of this fault will be 5 + 1: $\textbf{X}^*(t_s) = \textbf{X}(t_s^*),\ s > 8500$, where $t_s^* = \left[\frac{5(s - 8500)}{\omega - 8500} + 1\right]t_s$.
%
\item Fault 3B is a mutation of the underlying latent variable $t$ for the $z$ feature: $z^*(t_s) = z(\log t_s^*)$, $s \ge 8500$. This fault will dampen the underlying latent effect for $z$ if $t_s > 1$ and amplify this effect if $t_s < 1$.
%
\item Fault 3C is a polynomial mutation of the error for the $y$ feature in State 2 only and \emph{after} state rotation: $y^*(t_s) = y(t_s) + 2 * \textbf{e}_3(s) - 0.25$, for $s > 8500$.
\end{enumerate}

@@

This section provides a fully commented code walk-through for the main \texttt{msp.*} functions in the \texttt{mvMonitoring} package. This package depends on few other packages, and most of these dependencies are on commonly-used packages. The exception is \texttt{BMS}.
<<r load_mvMonitoring, warning=FALSE, message=FALSE>>=
library(devtools)
library(xts)
# install.packages("BMS") # If necessary
build("C:/Users/gabriel_odom/Documents/GitHub/mvMonitoring/mvMonitoring")
library(mvMonitoring)
@

\subsection{Generating Synthetic Data}
First, begin by generating multi-state data from a fault scenario. This code will yield observations under Fault 2A, as described in Section \ref{subsec:D_FaultDefns}. We choose the default options for the period length (7 days * 24 hours * 60 minutes = 10,080 observations), the starting index of the fault (8500 out of 10080), and the time stamp for beginning the data is 16 May of 2016 at 10:00AM (my wedding anniversary). As we can see from the structure of the output, \texttt{mspProcessData} returns an \texttt{xts} matrix with 10080 rows and four columns (the state indicator and the three features).

@@

\subsection{Train the Fault Detection Threshold}
Now that these observations are generated and stored in memory, the \texttt{mspTrain} function can train the MSAD-PCA model. The last 1620 observations (27 hours' worth) will be saved for validation. The \texttt{mspTrain} function takes in the training data matrix and corresponding class label column. If this function errors, make sure the label column is not included in the data matrix -- the constant values will cause a rank deficiency training covariance matrix. The \texttt{mspTrain} function will
\begin{enumerate}
\item Train on the first three days' worth of observations, as set by \texttt{trainObs}.
%
\item Scan the fourth day for anomalies, as set by \texttt{updateFreq}.
%
\item Remove any alarmed observations.
%
\item ``Forget'' the first day's observations.
%
\item ``Learn'' the non-alarmed observations from the fourth day.
%
\item Retrain and repeat until the end of the training matrix.
\end{enumerate}

Furthermore, the \texttt{Dynamic = TRUE} option means that the \texttt{mspTrain} function will include the lags specified by the \texttt{lagsIncluded} argument. Finally, the number of sequential anomalous observations necessary to raise an alarm is set at 5 by the \texttt{faultsToTriggerAlarm} argument. These last three arguments are set to their defaults.
<<r train>>=
train1A_xts <- fault1A_xts[1:8461,]
# This function will run in 13 seconds on the author's machine.
train1A_ls <- mspTrain(data = train1A_xts[,-1],
                       labelVector = train1A_xts[,1],
                       trainObs = 3 * 24 * 60,
                       updateFreq = 1 * 24 * 60,
                       Dynamic = TRUE,
                       lagsIncluded = 0:1,
                       faultsToTriggerAlarm = 5)
str(train1A_ls)
@
The \texttt{mspTrain} function returns a list of four objects:
\begin{enumerate}
\item \texttt{FaultChecks}: An \texttt{xts} matrix of monitoring statistics and associated indicators for all observations after the burn-in of \texttt{trainObs}. It will have 8461 - \texttt{trainObs} number of rows and five columns:
	\begin{enumerate}
	\item \texttt{SPE}: the SPE statistic for each observation.
    %
    \item \texttt{SPE\_Flag}: an indicator showing if the SPE statistic for that observation is beyond the calculated threshold; 0 is normal, 1 is flagged.
    %
    \item \texttt{T2}: the $T^2$ statistic for each observation.
    %
    \item \texttt{T2\_Flag}: an indicator showing if the $T^2$ statistic for that observation is beyond the calculated threshold; 0 is normal, 1 is flagged.
    %
    \item \texttt{Alarm}: an indicator showing if the observation is in a sequence of flagged observations; 0 is normal, 1 is alarmed.
	\end{enumerate}
%
\item \texttt{Non\_Alarmed\_Obs}: An \texttt{xts} matrix containing all observations with an alarm code of 0 from \texttt{FaultChecks}. Of note, this matrix contains the data, while the \texttt{FaultChecks} matrix only contains the monitoring statistics and indicators.
%
\item \texttt{Alarms}: An \texttt{xts} matrix of all the observations removed from the training data matrix.
%
\item \texttt{TrainingSpecs}: a list with length equal to the number of classes -- in this case 3. For each class, this list contains a list of six objects:
	\begin{enumerate}
	\item \texttt{SPE\_Threshold}: a named numeric scalar of the $1 - \alpha$ percentile of the non-parametric estimate of the SPE statistic density.
    %
    \item \texttt{T2\_Threshold}: a named numeric scalar of the $1 - \alpha$ percentile of the non-parametric estimate of the $T^2$ statistic density.
    %
    \item \texttt{projectionMatrix}: The $p \times q$ matrix of eigenvectors necessary to project a $p$-dimensional observation to $q$ dimensions. This is necessary to reduce the dimension of any test observation, and is used in calculating the SPE statistic for test observations.
    %
    \item \texttt{LambdaInv}: The inverse of the diagonal $q \times q$ matrix of eigenvalues. This matrix is used to calculate the $T^2$ statistic for test observations.
    %
    \item \texttt{muTrain}: The mean vector of the training observations. This is used to center the test observations on the training mean.
    %
    \item \texttt{RootPrecisTrain}: The $p \times p$ diagonal matrix of the inverse square roots of the feature variances. This is used to scale the test observations into the training scale.
	\end{enumerate}
\end{enumerate}

@@

\subsubsection{Monitoring the Test Data}
With the lagged test observations in the working environment, the \texttt{mspMonitor} function can be applied. This function (similarly to \texttt{mspTrain}) takes in the label information as a separate argument from the input data. Further, the \texttt{mspMonitor} function takes in the \texttt{TrainingSpecs} object returned in the results list from \texttt{mspTrain}. Notice that the first six rows of the matrix returned by \texttt{mspMonitor} are the exact same as the first six rows of the lagged test matrix, except that the rows of the matrix from the \texttt{mspMonitor} function have the monitoring statistics and corresponding indicator columns appended.
<<r monitor>>=
monitor1A_xts <- mspMonitor(observations = lagTest1A_xts[,-1],
                            labelVector = lagTest1A_xts[,1],
                            trainingSummary = train1A_ls$TrainingSpecs)
head(monitor1A_xts)
@

\subsection{Warn Operators During Alarms}
Note that all values in the \texttt{Alarm} column of the returned matrix above are recorded with \texttt{NA} values. This is because the \texttt{mspMonitor} function does not check the monitor output matrix for the sequential flags necessary to trigger an alarm. This is the responsibility of the \texttt{mspWarning} function. Because the \texttt{mspWarning} function is designed to test one incoming observation at a time through a script or batch file, the following example is designed to mimic the behavior of the \texttt{mspWarning} function as each new observation comes online.
<<r warnings>>=
alarm1A_xts <- monitor1A_xts
for(i in 1:nrow(alarm1A_xts)){
  if(i < (5 + 1)){
    alarm1A_xts[1:i,] <- mspWarning(alarm1A_xts[1:i,])
  }else{
    alarm1A_xts[(i - 5):i,] <- mspWarning(alarm1A_xts[(i - 5):i,])
  }
}
@
\noindent The fault was introduced at index 8500, which corresponds to about 40 minutes into the test hour.
<<r warning_Plot>>=
plot(alarm1A_xts[, ncol(alarm1A_xts)],
     main = "Alarm Codes for Test Data")
@
\noindent The alarm codes are
\begin{enumerate}
\item ``0'': No alarm.
\item ``1'': Hotelling's $T^2$ alarm.
\item ``2'': Squared Prediction Error alarm.
\item ``3'': Both alarms.
\end{enumerate}
\noindent As we can see, the monitoring function detects a process anomaly after 40 minutes into the test hour, and the warning function issues the corresponding alarms.
